---
title: "Project 1-Checkpoint2"
author: "Faran Igani and Tyler Stoen"
format:
  html:
    embed-resources: true
editor: visual
code-tools: true
execute:
  warning: false
  message: false
---

## Packages and Data

```{r}
#| label: Load packages 
library(tidyverse)
library(gt)

```

```{r}
#| label: Load the Data
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
```

```{r}
#Github URL
#https://github.com/FaranIgani/STAT-541---Project-1
```

## Checkpoint 2

-   **Two** data summaries that use existing (unmodified) columns in the data

-   **Two** data summaries that modify columns from the original data

**Table 1: Average plastic collected, number of events, and volunteers by year**

```{r}
#| label: Unmodified custom function
summarize_effort <- function(data) {
  data %>%
    group_by(year) %>%
    summarise(
        avg_grand_total = round(mean(grand_total, na.rm = TRUE), 3),
        avg_num_events = round(mean(num_events, na.rm = TRUE), 3),
        avg_volunteers = round(mean(volunteers, na.rm = TRUE), 3),
      .groups = "drop"
    )
}

summary_existing_1 <- summarize_effort(plastics)

colnames(summary_existing_1) <- c("Year", "Avg Total Plastic", "Avg Events", "Avg Volunteers")

```

```{r}
#| label: table1 w/ gt

table1 <- summary_existing_1 |>
  gt()

table1 <- table1 |>
  cols_label(
    Year = md("**Year**"),
    `Avg Total Plastic` = md("**Avg Total Plastic**"),
    `Avg Events` = md("**Avg Events**"),
    `Avg Volunteers` = md("**Avg Volunteers**")
  ) |>
  fmt_number(
    columns = c(`Avg Total Plastic`, `Avg Events`, `Avg Volunteers`),
    decimals = 2
  ) |>
  tab_header(
    title = md("**Table 1: Summary of Cleanup Effort by Year**")
  ) |>
  tab_source_note(
    source_note = "Average total plastic collected, number of events, and volunteers by year."
  )

table1
```

**Table 2: Dominant Plastic Type Recorded in Top 10 Waste-Producing Countries**

```{r}
#| label: Unmodified iteration
plastic_types <- c("hdpe", "ldpe", "o", "pet", "pp", "ps", "pvc")

dominant_plastic <- function(ctry) {
  df <- plastics %>%
    filter(country != "EMPTY") %>%
    mutate(country = str_to_title(country)) %>%   # match the casing
    filter(country == ctry) %>%
    summarise(across(all_of(plastic_types), ~ sum(.x, na.rm = TRUE)))
  
  tibble(
    country       = ctry,
    dominant_type = plastic_types[which.max(df)],
    dominant_n    = max(df)
  )
}

top_countries <- plastics %>%
  filter(country != "EMPTY") %>% 
  mutate(country = str_to_title(country)) %>% 
  group_by(country) %>%
  summarise(total = sum(grand_total, na.rm = TRUE)) %>%
  slice_max(total, n = 10) %>%
  pull(country)

top_countries %>%
  map(dominant_plastic) %>%
  list_rbind() %>%
  arrange(desc(dominant_n)) %>%
  mutate(dominant_type = dominant_type %>% 
           recode_values(
             "o" ~ "Other", 
             "ldpe" ~ "Low-Density Polyethylene",
             "hdpe" ~ "High-Density Polyethylene",
             "pet" ~ "Polyester",
             "pp" ~ "Polypropylene",
             "ps" ~ "Polystyrene",
             "pvc" ~ "PVC")
  ) %>% 
  gt() %>%
  tab_header(title = "Table 2: Dominant Plastic Type Recorded in Top 10 Waste-Producing Countries") %>%
  cols_label(
    country       = "Country",
    dominant_type = "Dominant Plastic Type",
    dominant_n    = "Amount of Dominant Plastic"
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>%
  fmt_number(columns = dominant_n, decimals = 0) %>%
  tab_source_note(
    source_note = "Dominant plastic type recorded among the top 10 waste-producing countries, based on the highest count of plastic items collected during cleanup events."
  )
```

**Table 3:** **Proportion of each plastic type by year**

```{r}
#| label: Modified custom function
plastic_props_year <- function(data) {
  
  data %>%
    filter(grand_total > 0, parent_company != "Grand Total") %>%
    group_by(year) %>%
    summarise(
      empty_prop = sum(empty, na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      hdpe_prop  = sum(hdpe,  na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      ldpe_prop  = sum(ldpe,  na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      other_prop = sum(o,     na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      pet_prop   = sum(pet,   na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      pp_prop    = sum(pp,    na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      ps_prop    = sum(ps,    na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      pvc_prop   = sum(pvc,   na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    rename(Year = year)
}

table3_year <- plastic_props_year(plastics)
```

```{r}
#| label: table3 w/ gt

table3 <- table3_year |>
  gt()

table3 <- table3 |>
  cols_label(
    Year = md("**Year**"),
    empty_prop = md("**Empty (%)**"),
    hdpe_prop  = md("**HDPE (%)**"),
    ldpe_prop  = md("**LDPE (%)**"),
    other_prop = md("**Other (%)**"),
    pet_prop   = md("**PET (%)**"),
    pp_prop    = md("**PP (%)**"),
    ps_prop    = md("**PS (%)**"),
    pvc_prop   = md("**PVC (%)**")
  ) |>
  fmt_percent(
    columns = c(empty_prop, hdpe_prop, ldpe_prop, other_prop,
                pet_prop, pp_prop, ps_prop, pvc_prop),
    decimals = 1
  ) |>
  tab_header(
    title = md("**Table 3: Proportion of Each Plastic Type by Year**")
  ) |>
  tab_source_note(
    source_note = "Percentages of each plastic type as a percentage of total plastic collected by year."
  )

table3

```

**Table 4: Parent Company that Contributes the Most Recorded Waste in Each Country**

```{r}
#| label: Modified iteration 
top_company_by_country <- function(ctry) {
  clean <- plastics %>% 
    filter(country == ctry) %>% 
    filter(!is.na(parent_company),
           !(parent_company %in% c("Grand Total", "null", "Unbranded")))
  
  country_total <- clean %>% 
  summarise(total = sum(grand_total, na.rm = TRUE)) %>% 
  pull(total)
  
  clean %>% 
    group_by(parent_company) %>% 
    summarise(total = sum(grand_total, na.rm = TRUE)) %>% 
    slice_max(total, n = 1) %>% 
    mutate(country = ctry,
           percent = total / country_total)
}

countries <- plastics %>% 
  filter(country != "EMPTY") %>% 
  distinct(country) %>%
  pull(country)

countries %>% 
  map(top_company_by_country) %>% 
  list_rbind() %>% 
  filter(!is.nan(percent)) %>%
  select(country, parent_company, percent) %>% 
  gt() %>% 
  cols_label(
    country        = "Country",
    parent_company = "Parent Company",
    percent        = "Percent"
  ) %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>% 
  fmt_percent(columns = percent, decimals = 2) %>% 
  tab_header(
    title = md("**Table 4: Company Contributing
                 the Most Recorded Waste by Country**")
  ) %>% 
  tab_source_note(
    source_note = "Parent company with the largest number of recorded waste items in each country." 
  )
```
