---
title: "Project 1-Checkpoint2"
author: "Faran Igani"
format:
  html:
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

## Packages and Data

```{r}
#| label: Load packages 
library(tidyverse)
library(gt)
```

```{r}
#| label: Load the Data
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
```

## Checkpoint 2

-   **Two** data summaries that use existing (unmodified) columns in the data

-   **Two** data summaries that modify columns from the original data

**Table 1: Average plastic collected, number of events, and volunteers by year**

```{r}
#| label: Unmodified custom function
summarize_effort <- function(data) {
  data %>%
    group_by(year) %>%
    summarise(
        avg_grand_total = round(mean(grand_total, na.rm = TRUE), 3),
        avg_num_events = round(mean(num_events, na.rm = TRUE), 3),
        avg_volunteers = round(mean(volunteers, na.rm = TRUE), 3),
      .groups = "drop"
    )
}

summary_existing_1 <- summarize_effort(plastics)

colnames(summary_existing_1) <- c("Year", "Avg Total Plastic", "Avg Events", "Avg Volunteers")

```

```{r}
#| label: table1 w/ gt

table1 <- summary_existing_1 |>
  gt()

table1 <- table1 |>
  cols_label(
    Year = md("**Year**"),
    `Avg Total Plastic` = md("**Avg Total Plastic**"),
    `Avg Events` = md("**Avg Events**"),
    `Avg Volunteers` = md("**Avg Volunteers**")
  ) |>
  fmt_number(
    columns = c(`Avg Total Plastic`, `Avg Events`, `Avg Volunteers`),
    decimals = 2
  ) |>
  tab_header(
    title = md("**Table 1: Summary of Cleanup Effort by Year**")
  ) |>
  tab_source_note(
    source_note = "Average total plastic collected, number of events, and volunteers by year."
  )

table1
```

**Table 2: Top 10 parent companies by grand total plastic within each year**

```{r}
#| label: Unmodified iteration
top_companies <- plastics %>%
  filter(
    parent_company != "Grand Total",
    !is.na(parent_company),
    parent_company != "null",
    parent_company != "NULL"
  ) %>%
  group_by(year, parent_company) %>%
  summarise(total_plastic = sum(grand_total, na.rm = TRUE), .groups = "drop") %>%
  arrange(year, desc(total_plastic)) %>%
  group_by(year) %>%
  slice_head(n = 10)

colnames(top_companies) <- c( "Year" ,"Parent Company","Total Plastic")

# with cable
# knitr::kable(
#   summary_existing_1,
#   caption = "Table 2: Top contributing parent companies ranked by total plastic collected within each year."
# )
```

```{r}
#| label: table2 w/ gt

table2 <- top_companies |>
  gt()

table2 <- table2 |>
  cols_label(
    Year = md("**Year**"),
    `Parent Company` = md("**Parent Company**"),
    `Total Plastic` = md("**Total Plastic**")
  ) |>
  fmt_number(
    columns = c(`Total Plastic`),
    decimals = 0
  ) |>
  tab_header(
    title = md("**Table 2: Top Parent Companies by Total Plastic Collected**")
  ) |>
  tab_source_note(
    source_note = "Top 10 parent companies by total plastic collected within each year, excluding missing and aggregate entries."
  )

table2
```

**Table 3:** **Proportion of each plastic type by year**

```{r}
#| label: Modified custom function
plastic_props_year <- function(data) {
  
  data %>%
    filter(grand_total > 0, parent_company != "Grand Total") %>%
    group_by(year) %>%
    summarise(
      empty_prop = sum(empty, na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      hdpe_prop  = sum(hdpe,  na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      ldpe_prop  = sum(ldpe,  na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      other_prop = sum(o,     na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      pet_prop   = sum(pet,   na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      pp_prop    = sum(pp,    na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      ps_prop    = sum(ps,    na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      pvc_prop   = sum(pvc,   na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    rename(Year = year)
}

table3_year <- plastic_props_year(plastics)
```

```{r}
#| label: table3 w/ gt

table3 <- table3_year |>
  gt()

table3 <- table3 |>
  cols_label(
    Year = md("**Year**"),
    empty_prop = md("**Empty (%)**"),
    hdpe_prop  = md("**HDPE (%)**"),
    ldpe_prop  = md("**LDPE (%)**"),
    other_prop = md("**Other (%)**"),
    pet_prop   = md("**PET (%)**"),
    pp_prop    = md("**PP (%)**"),
    ps_prop    = md("**PS (%)**"),
    pvc_prop   = md("**PVC (%)**")
  ) |>
  fmt_percent(
    columns = c(empty_prop, hdpe_prop, ldpe_prop, other_prop,
                pet_prop, pp_prop, ps_prop, pvc_prop),
    decimals = 1
  ) |>
  tab_header(
    title = md("**Table 3: Proportion of Each Plastic Type by Year**")
  ) |>
  tab_source_note(
    source_note = "Percentages of each plastic type as a percentage of total plastic collected, by year."
  )

table3

```

**Table 4: Total plastic counts by type and year**

```{r}
#| label: Modified iteration

plastic_long <- plastics %>%
  pivot_longer(
    cols = c(empty, hdpe, ldpe, o, pet, pp, ps, pvc),
    names_to = "plastic_type",
    values_to = "count"
  )

years <- sort(unique(plastics$year))

plastic_summary <- map_dfr(years, function(y) {
  plastic_long %>%
    filter(year == y, parent_company != "Grand Total") %>%
    group_by(plastic_type) %>%
    summarise(total_count = sum(count, na.rm = TRUE), .groups = "drop") %>%
    mutate(year = y)%>%
    arrange(desc(total_count))
})


colnames(plastic_summary) <- c("Plastic Type", "Total Count","Year")

```

```{r}
#| label: table4 w/ gt

plastic_summary <- plastic_summary %>%
  select(Year, `Plastic Type`, `Total Count`)

table4 <- plastic_summary |>
  gt()

table4 <- table4 |>
  cols_label(
    `Plastic Type` = md("**Plastic Type**"),
    `Total Count` = md("**Total Count**"),
    Year = md("**Year**")
  ) |>
  fmt_number(
    columns = c(`Total Count`),
    decimals = 0
  ) |>
  tab_header(
    title = md("**Table 4: Total Plastic Counts by Plastic Type and Year**")
  ) |>
  tab_source_note(
    source_note = "Total plastic counts by plastic type and year after reshaping the data into long format."
  )

table4

```
