---
title: "Project 1 Checkpoint 3"
author: "Faran Igani and Tyler Stoen"
format:
  html:
    embed-resources: true
editor: visual
code-tools: true
execute:
  warning: false
  message: false
---

## Packages and Data

```{r}
#| label: Load packages 
library(tidyverse)
library(gt)
library(rvest)
library(ggplot2)
library(dplyr)
library(httr2)
library(jsonlite)
```

```{r}
#| label: Load the Data
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
```

```{r}
#Github URL
#https://github.com/FaranIgani/STAT-541---Project-1
```

## Checkpoint 3

-   **two** new country-level data sources, One acquired using APIs, the other using webscraping.

-   descriptions of each of the new datasets

-   a “meta” dataset containing **all** the acquired data (joined together)

-   **at least two** data summaries OR visualizations incorporating the additional country-level data

## New data

```{r}
#| label: API, World Bank  API

base_url <- "https://api.worldbank.org/v2/country/all/indicator/"

pop_code <- "SP.POP.TOTL"
gdp_code <- "NY.GDP.MKTP.CD"

pop_url <- str_c(base_url, pop_code, "?format=json&per_page=20000")
gdp_url <- str_c(base_url, gdp_code, "?format=json&per_page=20000")
```

```{r}
#| label: request
pop_raw <- request(pop_url) |>
  req_perform() |>
  resp_body_string() |>
  fromJSON(flatten = TRUE)

gdp_raw <- request(gdp_url) |>
  req_perform() |>
  resp_body_string() |>
  fromJSON(flatten = TRUE)
```

```{r}
#| label: get data and combine
pop_data <- pop_raw[[2]] |>
  as_tibble() |>
  transmute(
    country = country.value,
    year = as.numeric(date),
    population = as.numeric(value)
  ) |>
  filter(year %in% c(2019, 2020))

gdp_data <- gdp_raw[[2]] |>
  as_tibble() |>
  transmute(
    country = country.value,
    year = as.numeric(date),
    gdp = as.numeric(value)
  ) |>
  filter(year %in% c(2019, 2020))

wb_data <- pop_data |>
  inner_join(gdp_data, by = c("country", "year"))
```

The first supplemental dataset was obtained using the World Bank API through the `WDI` R package. This dataset includes country-level indicators such as total population and gross domestic product (GDP) for the years 2019 and 2020. These variables provide measures of country size and economic development, which can be used to examine whether plastic waste generation is associated with population or economic activity. The dataset is structured at the country-year level, allowing it to be merged directly with the plastics dataset using country and year as keys.

```{r}
#| label: Webscraping, Wiki population
#not multi page
url <- "https://en.wikipedia.org/wiki/List_of_countries_by_average_yearly_temperature"

temp_data <- read_html(url) %>%
  html_table(fill = TRUE)

temp_data <- temp_data[[1]]
```

```{r}
#| label: Webscraping, Wiki population density

scrape_density <- function(country_name, wiki_name) {
  
  url <- str_c("https://en.wikipedia.org/wiki/", wiki_name)
  
  out <- tryCatch({
    page <- read_html(url)
    
    rows <- page %>%
      html_elements("table.infobox tr")
    
    row_text <- rows %>% html_text2()
    
    density_idx <- str_which(row_text, regex("density", ignore_case = TRUE))
    
    if (length(density_idx) == 0) {
      return(tibble(country = country_name, pop_density = NA_real_))
    }
    
    density_row <- rows[density_idx[1]]
    
    cells <- density_row %>% html_elements("th, td") %>% html_text2()
    
    if (length(cells) < 2) {
      return(tibble(country = country_name, pop_density = NA_real_))
    }
    
    value <- parse_number(cells[length(cells)])
    
    tibble(country = country_name, pop_density = value)
    
  }, error = function(e) {
    tibble(country = country_name, pop_density = NA_real_)
  })
  
  out
}
```

```{r}
#| label: mapping to countries in plastic data

#normalize names
plastics_clean <- plastics %>%
  mutate(country = str_to_lower(country)) %>%
  mutate(country = str_replace_all(country, "_", " ")) %>%
  mutate(country = str_replace_all(country, "[()&]", "")) %>%
  mutate(country = str_squish(country)) %>%
  filter(country != "empty")


countries <- plastics_clean %>%
  distinct(country) %>%
  mutate(
    wiki_country = case_when(
      country == "united states of america" ~ "United_States",
      country == "hong kong" ~ "Hong_Kong",
      country == "south africa" ~ "South_Africa",
      country == "sri lanka" ~ "Sri_Lanka",
      country == "cote divoire" ~ "Ivory_Coast",
      country == "taiwan republic of china roc" ~ "Taiwan",
      country == "taiwan_republic of china roc" ~ "Taiwan",
      country == "united kingdom of great britain northern ireland" ~ "United_Kingdom",
      country == "united kingdom of great britain & northern ireland" ~ "United_Kingdom",
      country == "korea" ~ "South_Korea",
      country == "cyprus" ~ "Cyprus",
      country == "ecuador" ~ "Ecuador",
      TRUE ~ str_replace_all(str_to_title(country), " ", "_")
    )
  )

density_data <- map2_dfr(
  countries$country,
  countries$wiki_country,
  scrape_density
)
```

```{r}
density_data %>%
  arrange(desc(pop_density)) 
```

The web-scraped country-level dataset was collected from multiple Wikipedia country pages using iteration. For each country in the plastics dataset, a country-specific Wikipedia URL was constructed and scraped with rvest package. Population density values were extracted from each page’s infobox, then combined into a single country-level dataset. This approach satisfies the requirement of gathering data from a multi-page source using iteration.

## Clean

```{r}
#| label: temp clean
names(temp_data) <- make.names(names(temp_data), unique = TRUE)

temp_data <- temp_data %>%
  rename(country = `Countryor.region`) %>%
  select(country, Temperature)

```

```{r}
temp_data <- temp_data %>%
  mutate(
    Temperature = readr::parse_number(Temperature)
  )
```

```{r}
#| label: align names

clean_names_fn <- function(df) {
  df %>%
  # mutate(country = str_to_lower(country)) %>%
    mutate(country = str_replace_all(country, "[[:punct:]]", "")) %>%
    mutate(country = str_squish(country))
}

plastics_clean <- clean_names_fn(plastics)
temp_clean     <- clean_names_fn(temp_data)
wb_clean       <- clean_names_fn(wb_data)

density_data <- density_data %>%
  mutate(country = str_to_title(country))
```

```{r}
#| label: year to numeric
plastics_clean <- plastics_clean %>%
  mutate(year = as.numeric(year))

wb_clean <- wb_clean %>%
  mutate(year = as.numeric(year))
```

## Combined Data-sets

```{r}
#| label: merge plastic with wb
plastics_wb <- plastics_clean %>%
  inner_join(wb_clean, by = c("country", "year"))
plastics_wb
```

```{r}
#| label: wb with temp
plastics_full <- plastics_wb %>%
  inner_join(temp_clean, by = "country")
plastics_full
```

```{r}
#| label: Full with pop density

plastics_full <- plastics_full %>%
  inner_join(density_data, by = "country")
plastics_full
```

## Two Summaries/Vis of new data

```{r}
#| label: plastic per person for each country
plastic_per_capita <- plastics_full %>%
  group_by(country, year) %>%
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    plastic_per_person = total_plastic / population,
    .groups = "drop"
  ) %>%
  arrange(desc(plastic_per_person))%>%
  slice_head(n = 20)


table_per_capita <- plastic_per_capita |>
  gt()

table_per_capita <- table_per_capita |>
  cols_label(
    country = md("**Country**"),
    year = md("**Year**"),
    total_plastic = md("**Total Plastic**"),
    population = md("**Population**"),
    plastic_per_person = md("**Plastic per Person**")
  ) |>
  fmt_number(
    columns = c(total_plastic, population),
    decimals = 0
  ) |>
  fmt_number(
    columns = plastic_per_person,
    decimals = 5
  ) |>
  tab_header(
    title = md("**Top 20 Countries by Plastic Contributed per Person**")
  ) |>
  tab_source_note(
    source_note = "Plastic per person was calculated as total plastic collected divided by population."
  )

table_per_capita
```

```{r}
#| label: Plastic vs GDP
country_summary <- plastics_full %>%
  group_by(country, year) %>%
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    gdp = mean(gdp, na.rm = TRUE),
    .groups = "drop"
  )

country_summary_filtered <- country_summary %>%
  filter(total_plastic < quantile(total_plastic, 0.99, na.rm = TRUE))

ggplot(country_summary_filtered, aes(x = gdp, y = total_plastic)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  scale_x_continuous(labels = scales::label_number(scale = 1e-9, suffix = "B")) +
  labs(
    title = "Total Plastic vs GDP by Country",
    x = "GDP (Billions USD)",
    y = "Total Plastic Collected"
  ) +
  theme_minimal()
```

```{r}
#| label: Plastic vs Temp
temp_summary <- plastics_full %>%
  group_by(country, year) %>%
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    Temperature = mean(Temperature, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(temp_summary, aes(x = Temperature, y = total_plastic)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  labs(
    title = "Total Plastic vs Average Temperature by Country",
    x = "Average Temperature (°C)",
    y = "Total Plastic Collected"
  ) +
  theme_minimal()
```

```{r}
density_summary <- plastics_full %>%
  group_by(country, year) %>%
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    pop_density = mean(pop_density, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(pop_density < quantile(pop_density, 0.99, na.rm = TRUE))

ggplot(density_summary, aes(x = pop_density, y = total_plastic)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  labs(
    title = "Total Plastic vs Population Density by Country",
    x = "Population Density",
    y = "Total Plastic Collected"
  ) +
  theme_minimal()
```
