---
title: "Project 1 Checkpoint 3"
author: "Faran Igani and Tyler Stoen"
format:
  html:
    embed-resources: true
editor: visual
code-tools: true
execute:
  warning: false
  message: false
---

## Packages and Data

```{r}
#| label: Load packages 
library(tidyverse)
library(gt)
library(rvest)
library(ggplot2)
library(dplyr)
library(httr2)
library(jsonlite)
library(httr)
library(glue)
```

```{r}
#| label: Load the Data
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
```

```{r}
#Github URL
#https://github.com/FaranIgani/STAT-541---Project-1
```

## Checkpoint 3

-   **two** new country-level data sources, One acquired using APIs, the other using webscraping.

-   descriptions of each of the new datasets

-   a “meta” dataset containing **all** the acquired data (joined together)

-   **at least two** data summaries OR visualizations incorporating the additional country-level data

## Data from API

### API 1: restcountries.com

```{r}
#| label: data-from-api

base_url <- "https://restcountries.com/v4/"

restcountries <- GET(glue("{base_url}all?fields=name,continents,region,subregion,unMember,gini,density,geolocation"))

restcountries

restcountries_tidy <- restcountries %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>%
  as_tibble()

```

```{r}
#| label: merge-plastics

library(stringi)

normalize_name <- function(x) {
  x %>% 
    stri_trans_general("Latin-ASCII") %>% 
    str_to_lower() %>% 
    str_remove_all("[^a-z ]") %>% 
    str_trim() %>% 
    str_squish()
}

plastics_clean <- plastics %>% 
  filter(country != "EMPTY") %>% 
  mutate(
    country = case_when(
      country == "Korea" ~ "South Korea",
      country == "Taiwan_ Republic of China (ROC)" ~ "Taiwan",
      country == "United Kingdom of Great Britain & Northern Ireland" ~ 
        "United Kingdom",
      TRUE ~ country
    )
  ) %>%
  mutate(country_key = normalize_name(country))

restcountries_clean <- restcountries_tidy %>% 
  select(name.common, name.official, region, subregion,
         density, unMember, geolocation.latitude, 
         geolocation.longitude) %>% 
  mutate(common_key = recode(
    normalize_name(name.common),
    "ivory coast" = "cote divoire"
    ),
         official_key = normalize_name(name.official))
  

# Different name types, try both

joined <- plastics_clean %>%
  left_join(
    restcountries_clean %>% select(-official_key),
    by = c("country_key" = "common_key")
  )

matched <- joined %>% filter(!is.na(region))
unmatched <- joined %>% filter(is.na(region)) %>%
  select(names(plastics_clean))

refilled <- unmatched %>%
  left_join(
    restcountries_clean %>% select(-common_key),
    by = c("country_key" = "official_key")
  )

joined <- bind_rows(matched, refilled)
```

#### Visual 1

```{r}
#| label: total plastic waste per cont
joined %>%
  filter(!is.na(region)) %>% 
  group_by(country, region) %>%
  summarize(
    total = sum(grand_total, na.rm = TRUE), .groups = "drop"
  ) %>% 
  ggplot(aes(x=total, y=region, colour=region)) +
  geom_boxplot() +
  labs(
    title = "Total Plastic Waste by Region",
    x= "Total Plastic Waste",
    y= "Region"
  ) +
  scale_color_brewer(palette = "Pastel1", guide = "none") +
  theme_minimal()
```

### API 2: worldbank

```{r}
#| label: API, World Bank  API

base_url <- "https://api.worldbank.org/v2/country/all/indicator/"

pop_code <- "SP.POP.TOTL"
gdp_code <- "NY.GDP.MKTP.CD"

pop_url <- str_c(base_url, pop_code, "?format=json&per_page=20000")
gdp_url <- str_c(base_url, gdp_code, "?format=json&per_page=20000")

pop_raw <- GET(pop_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>%
  pluck(2) %>% 
  as_tibble() %>% 
  select(
    country.value,
    year = date,
    population = value
  ) %>% 
  mutate(year = as.numeric(year)) %>% 
  filter(year %in% c(2019, 2020))

gdp_raw <- GET(gdp_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>% 
  pluck(2) %>% 
  as_tibble() %>% 
  select(
    country.value,
    year = date,
    gdp = value
  ) %>% 
  mutate(year = as.numeric(year)) %>% 
  filter(year %in% c(2019, 2020))

pop_and_gdp_raw <- inner_join(pop_raw, gdp_raw, by = c("country.value", "year")) %>% 
  mutate(country.value = recode(
           country.value,
          "Turkiye" = "Turkey",
          "Hong Kong SAR, China" = "Hong Kong",
          "Viet Nam" = "Vietnam",
          "Korea, Rep." = "South Korea",
          "United States" = "United States of America"
          ),
         country_code = normalize_name(country.value))

joined2 <- joined %>% 
  left_join(pop_and_gdp_raw, by = c("country_key" = "country_code", "year")) %>% 
  mutate(
    gdppercapita = gdp / population
  )
                              
```

The first supplemental dataset was obtained using the World Bank API through the `WDI` R package. This dataset includes country-level indicators such as total population and gross domestic product (GDP) for the years 2019 and 2020. These variables provide measures of country size and economic development, which can be used to examine whether plastic waste generation is associated with population or economic activity. The dataset is structured at the country-year level, allowing it to be merged directly with the plastics dataset using country and year as keys.

#### Visual 2

```{r}
#| label: Plastic waste vs GDP
joined2 %>% 
  group_by(country, gdppercapita, region) %>% 
  summarize(
    total = sum(grand_total, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = gdppercapita, y = total, colour = region)) +
  geom_point() +
  labs(
    title = "Plastic Waste Per Capita",
    x= "GDP per Capita",
    y= "Total Recorded Plastic Waste"
  ) +
  theme_minimal()
  
```

#### Table 1

```{r}
#| label: Plaste waste per person
joined2 %>%
  group_by(country, year) %>%
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    plastic_per_person = total_plastic / population,
    .groups = "drop"
  ) %>%
  arrange(desc(plastic_per_person))%>%
  slice_head(n = 20) %>% 
  gt() %>% 
  cols_label(
    country = md("**Country**"),
    year = md("**Year**"),
    total_plastic = md("**Total Plastic**"),
    population = md("**Population**"),
    plastic_per_person = md("**Plastic per Person**")
  ) %>% 
  fmt_number(
    columns = c(total_plastic, population),
    decimals = 0
  ) %>% 
  fmt_number(
    columns = plastic_per_person,
    decimals = 5
  ) %>% 
  tab_header(
    title = md("**Top 20 Countries by Plastic Contributed per Person**")
  ) %>% 
  tab_source_note(
    source_note = "Plastic per person was calculated as total plastic collected divided by population."
  )


```

## Web-Scraped Data

```{r}
#| label: Webscraping, Wiki population
#not multi page
url <- "https://en.wikipedia.org/wiki/List_of_countries_by_average_yearly_temperature"

temp_data <- read_html(url) %>%
  html_table(fill = TRUE)

temp_data <- temp_data[[1]]

names(temp_data) <- make.names(names(temp_data), unique = TRUE)

temp_data <- temp_data %>% 
  rename(country = `Countryor.region`) %>%
  mutate(
    Temperature = readr::parse_number(Temperature),
    country_key = normalize_name(country)
  ) %>% 
  select(country_key, Temperature)
```

```{r}
#| label: Webscraping, Wiki population density multi page

scrape_density <- function(country_name, wiki_name) {
  
  url <- str_c("https://en.wikipedia.org/wiki/", wiki_name)
  
  out <- tryCatch({
    page <- read_html(url)
    
    rows <- page %>%
      html_elements("table.infobox tr")
    
    row_text <- rows %>% html_text2()
    
    density_idx <- str_which(row_text, regex("density", ignore_case = TRUE))
    
    if (length(density_idx) == 0) {
      return(tibble(country_key = country_name, pop_density = NA_real_))
    }
    
    density_row <- rows[density_idx[1]]
    
    cells <- density_row %>% html_elements("th, td") %>% html_text2()
    
    if (length(cells) < 2) {
      return(tibble(country_key = country_name, pop_density = NA_real_))
    }
    
    value <- parse_number(cells[length(cells)])
    
    tibble(country_key = country_name, pop_density = value)
    
  }, error = function(e) {
    tibble(country_key = country_name, pop_density = NA_real_)
  })
  
  out
}
```

```{r}
#| label: mapping to countries in plastic data

wiki_countries <- joined2 %>%
  distinct(country_key) %>%
  mutate(
    wiki_country = recode(country_key,
      "united states of america" = "United_States",
      "hong kong" = "Hong_Kong",
      "south africa" = "South_Africa",
      "sri lanka" = "Sri_Lanka",
      "cote divoire" = "Ivory_Coast",
      "taiwan republic of china roc" = "Taiwan",
      "united kingdom of great britain northern ireland" = "United_Kingdom",
      "korea" = "South_Korea",
      .default = str_replace_all(str_to_title(country_key), " ", "_")
    )
  )

density_data <- map2_dfr(
  wiki_countries$country_key,
  wiki_countries$wiki_country,
  scrape_density
)
```

```{r}
joined_all <- joined2 %>% 
  left_join(temp_data, by = "country_key") %>% 
  left_join(density_data, by = "country_key")
```

The web-scraped country-level dataset was collected from multiple Wikipedia country pages using iteration. For each country in the plastics dataset, a country-specific Wikipedia URL was constructed and scraped with rvest package. Population density values were extracted from each page’s infobox, then combined into a single country-level dataset. This approach satisfies the requirement of gathering data from a multi-page source using iteration.

#### Visual 3

```{r}
#| label: Total lastic vs temp
joined_all %>%
  group_by(country, year) %>%
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    Temperature = mean(Temperature, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = Temperature, y = total_plastic)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
    labs(
      title = "Total Plastic vs Average Temperature by Country",
      x = "Average Temperature (°C)",
      y = "Total Plastic Collected"
    ) +
    theme_minimal()
```

#### Visual 4

```{r}
#| label: Total lastic vs population density 
density_summary <- joined_all %>%
  group_by(country, year) %>%
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    pop_density = mean(pop_density, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(pop_density < quantile(pop_density, 0.99, na.rm = TRUE))

ggplot(density_summary, aes(x = pop_density, y = total_plastic)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  labs(
    title = "Total Plastic vs Population Density by Country",
    x = "Population Density",
    y = "Total Plastic Collected"
  ) +
  theme_minimal()
```
