---
title: "Project 1 Checkpoint 4"
author: "Faran Igani and Tyler Stoen"
format:
  html:
    embed-resources: true
editor: visual
code-tools: true
execute:
  warning: false
  message: false
---

## Packages and Data

```{r}
#| label: Load packages 
library(tidyverse)
library(gt)
library(rvest)
library(ggplot2)
library(dplyr)
library(httr2)
library(jsonlite)
library(httr)
library(glue)
```

```{r}
#| label: Load the Data
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
```

## Github URL

<https://github.com/FaranIgani/STAT-541---Project-1>

## Checkpoint 4

-   **two** professionally formatted visualizations

    -   every visualization **must** have alternative (alt) text, non-default colors, a non-default theme, and clear labels

    -   **at least one** visualization must use a more complex geometry (e.g., `geom_segment()`, `geom_density_ridge()` `geom_ribbon()`, `geom_tile()`, `geom_hex()`)

-   fits **one** statistical model addressing one of the research questions posed

-   professionally formats the output of that model to communicate the results

    -   Formatted values (e.g., decimals, percentages, currencies)

    -   Boldface headers

    -   Captions

## Data from API

### API 1: restcountries.com

```{r}
#| label: data-from-api

base_url <- "https://restcountries.com/v4/"

restcountries <- GET(glue("{base_url}all?fields=name,continents,region,subregion,unMember,gini,density,geolocation"))

restcountries

restcountries_tidy <- restcountries %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>%
  as_tibble()

```

```{r}
#| label: merge-plastics

library(stringi)

normalize_name <- function(x) {
  x %>% 
    stri_trans_general("Latin-ASCII") %>% 
    str_to_lower() %>% 
    str_remove_all("[^a-z ]") %>% 
    str_trim() %>% 
    str_squish()
}

plastics_clean <- plastics %>% 
  filter(country != "EMPTY") %>% 
  mutate(
    country = case_when(
      country == "Korea" ~ "South Korea",
      country == "Taiwan_ Republic of China (ROC)" ~ "Taiwan",
      country == "United Kingdom of Great Britain & Northern Ireland" ~ 
        "United Kingdom",
      TRUE ~ country
    )
  ) %>%
  mutate(country_key = normalize_name(country))

restcountries_clean <- restcountries_tidy %>% 
  select(name.common, name.official, region, subregion,
         density, unMember, geolocation.latitude, 
         geolocation.longitude) %>% 
  mutate(common_key = recode(
    normalize_name(name.common),
    "ivory coast" = "cote divoire"
    ),
         official_key = normalize_name(name.official))
  

# Different name types, try both

joined <- plastics_clean %>%
  left_join(
    restcountries_clean %>% select(-official_key),
    by = c("country_key" = "common_key")
  )

matched <- joined %>% filter(!is.na(region))
unmatched <- joined %>% filter(is.na(region)) %>%
  select(names(plastics_clean))

refilled <- unmatched %>%
  left_join(
    restcountries_clean %>% select(-common_key),
    by = c("country_key" = "official_key")
  )

joined <- bind_rows(matched, refilled)
```

### API 2: worldbank

```{r}
#| label: API, World Bank  API

base_url <- "https://api.worldbank.org/v2/country/all/indicator/"

pop_code <- "SP.POP.TOTL"
gdp_code <- "NY.GDP.MKTP.CD"

pop_url <- str_c(base_url, pop_code, "?format=json&per_page=20000")
gdp_url <- str_c(base_url, gdp_code, "?format=json&per_page=20000")

pop_raw <- GET(pop_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>%
  pluck(2) %>% 
  as_tibble() %>% 
  select(
    country.value,
    year = date,
    population = value
  ) %>% 
  mutate(year = as.numeric(year)) %>% 
  filter(year %in% c(2019, 2020))

gdp_raw <- GET(gdp_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>% 
  pluck(2) %>% 
  as_tibble() %>% 
  select(
    country.value,
    year = date,
    gdp = value
  ) %>% 
  mutate(year = as.numeric(year)) %>% 
  filter(year %in% c(2019, 2020))

pop_and_gdp_raw <- inner_join(pop_raw, gdp_raw, by = c("country.value", "year")) %>% 
  mutate(country.value = recode(
           country.value,
          "Turkiye" = "Turkey",
          "Hong Kong SAR, China" = "Hong Kong",
          "Viet Nam" = "Vietnam",
          "Korea, Rep." = "South Korea",
          "United States" = "United States of America"
          ),
         country_code = normalize_name(country.value))

joined2 <- joined %>% 
  left_join(pop_and_gdp_raw, by = c("country_key" = "country_code", "year")) %>% 
  mutate(
    gdppercapita = gdp / population
  )
                              
```

## Web-Scraped Data

```{r}
#| label: Webscraping, Wiki population
#not multi page
url <- "https://en.wikipedia.org/wiki/List_of_countries_by_average_yearly_temperature"

temp_data <- read_html(url) %>%
  html_table(fill = TRUE)

temp_data <- temp_data[[1]]

names(temp_data) <- make.names(names(temp_data), unique = TRUE)

temp_data <- temp_data %>% 
  rename(country = `Countryor.region`) %>%
  mutate(
    Temperature = readr::parse_number(Temperature),
    country_key = normalize_name(country)
  ) %>% 
  select(country_key, Temperature)
```

```{r}
#| label: Webscraping, Wiki population density multi page

scrape_density <- function(country_name, wiki_name) {
  
  url <- str_c("https://en.wikipedia.org/wiki/", wiki_name)
  
  out <- tryCatch({
    page <- read_html(url)
    
    rows <- page %>%
      html_elements("table.infobox tr")
    
    row_text <- rows %>% html_text2()
    
    density_idx <- str_which(row_text, regex("density", ignore_case = TRUE))
    
    if (length(density_idx) == 0) {
      return(tibble(country_key = country_name, pop_density = NA_real_))
    }
    
    density_row <- rows[density_idx[1]]
    
    cells <- density_row %>% html_elements("th, td") %>% html_text2()
    
    if (length(cells) < 2) {
      return(tibble(country_key = country_name, pop_density = NA_real_))
    }
    
    value <- parse_number(cells[length(cells)])
    
    tibble(country_key = country_name, pop_density = value)
    
  }, error = function(e) {
    tibble(country_key = country_name, pop_density = NA_real_)
  })
  
  out
}
```

```{r}
#| label: mapping to countries in plastic data

wiki_countries <- joined2 %>%
  distinct(country_key) %>%
  mutate(
    wiki_country = recode(country_key,
      "united states of america" = "United_States",
      "hong kong" = "Hong_Kong",
      "south africa" = "South_Africa",
      "sri lanka" = "Sri_Lanka",
      "cote divoire" = "Ivory_Coast",
      "taiwan republic of china roc" = "Taiwan",
      "united kingdom of great britain northern ireland" = "United_Kingdom",
      "korea" = "South_Korea",
      .default = str_replace_all(str_to_title(country_key), " ", "_")
    )
  )

density_data <- map2_dfr(
  wiki_countries$country_key,
  wiki_countries$wiki_country,
  scrape_density
)
```

```{r}
joined_all <- joined2 %>% 
  left_join(temp_data, by = "country_key") %>% 
  left_join(density_data, by = "country_key")
```

## 2 Visuals 1 Model Fit:

```{r}
library(tidyverse)

country_year <- joined_all %>%
  group_by(country, year, region) %>%
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    gdp = mean(gdp, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    Temperature = mean(Temperature, na.rm = TRUE),
    pop_density = mean(pop_density, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(
    !is.na(total_plastic),
    !is.na(gdp),
    !is.na(Temperature),
    !is.na(region)
  )
```

**Visual 1**

```{r}
library(ggplot2)
library(hexbin)
library(scales)
#| fig-cap: "Hexbin plot showing the relationship between GDP and total plastic collected across country-year observations. Darker hexagons indicate more observations concentrated in that area. Most countries cluster at lower GDP values, with plastic totals varying widely."

ggplot(country_year, aes(x = gdp, y = total_plastic)) +
  geom_hex(bins = 20) +
  scale_x_log10(labels = label_number(scale = 1e-9, suffix = "B")) +
  scale_fill_gradient(low = "#cfe8f3", high = "#045a8d", name = "Count") +
  labs(
    title = "Plastic Waste and GDP by Country",
    subtitle = "Each hexagon shows the concentration of country-year observations",
    x = "GDP (Billions USD, log scale)",
    y = "Total Plastic Waste"
  ) +
  theme_light(base_size = 15)
```

**Visual 2**

```{r}
#| fig-cap: "Heatmap showing average plastic collected by region in 2019 and 2020. Darker cells indicate regions with higher average plastic totals."
region_year <- country_year %>%
  group_by(region, year) %>%
  summarise(
    avg_plastic = mean(total_plastic, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(region_year, aes(x = factor(year), y = region, fill = avg_plastic)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient(low = "#fee5d9", high = "#de2d26", name = "Avg Plastic") +
  labs(
    title = "Average Plastic Collected by Region and Year",
    x = "Year",
    y = "Region"
  ) +
  theme_minimal(base_size = 15)
```

**Visual 3**

```{r}
#| label: Total plastic vs population density
#| fig-cap: "Scatterplot showing the relationship between population density and total plastic collected across country-year observations. The fitted regression line summarizes the overall trend after removing extreme population density outliers."


density_summary <- joined_all %>%
  group_by(country, year) %>%
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    pop_density = mean(pop_density, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(pop_density < quantile(pop_density, 0.99, na.rm = TRUE))

ggplot(density_summary, aes(x = pop_density, y = total_plastic)) +
  geom_point(
    color = "#c44e52",
    alpha = 0.7,
    size = 2
  ) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = TRUE,
    color = "#7f0000",
    linewidth = 1
  ) +
  scale_x_log10() +
  labs(
    title = "Total Plastic Collected vs Population Density",
    subtitle = "Country-year observations with extreme population density outliers removed",
    x = "Population Density (log scale)",
    y = "Total Plastic Collected"
  ) +
  theme_light(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

**Visual** 4

```{r}
#| label: Total plastic waste by region
#| fig-cap: "Boxplot showing the distribution of total plastic collected across regions. Each box represents the spread of country-level totals within a region, highlighting variation in plastic waste across different parts of the world."

joined_all %>%
  filter(!is.na(region)) %>% 
  filter(!is.na(region), region != "Oceania") %>% 
  group_by(country, region) %>%
  summarise(
    total = sum(grand_total, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(region = fct_reorder(region, total, .fun = median)) %>%
  ggplot(aes(x = total, y = region, fill = region)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.4) +
  scale_fill_brewer(palette = "Reds", guide = "none") +
  labs(
    title = "Distribution of Total Plastic Waste by Region",
    subtitle = "Country-level totals aggregated within each region",
    x = "Total Plastic Collected",
    y = "Region"
  ) +
  theme_light(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

**Visual 4.5**

```{r}
#| label: Total plastic waste by region (ridge)
#| fig-cap: "Density ridge plot showing the distribution of total plastic collected across regions. Each ridge represents the distribution of country-level plastic totals within a region."

joined_all %>%
  filter(!is.na(region)) %>% 
  filter(!is.na(region), region != "Oceania") %>% 
  group_by(country, region) %>%
  summarise(
    total = sum(grand_total, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(region = fct_reorder(region, total, .fun = median)) %>%
  ggplot(aes(x = total, y = region, fill = region)) +
  geom_density_ridges(
    alpha = 0.8,
    scale = 1.2,
    color = "white",
    rel_min_height = 0.01
  ) +
  scale_fill_brewer(palette = "Reds", guide = "none") +
  labs(
    title = "Distribution of Total Plastic Waste by Region",
    subtitle = "Density of country-level totals within each region",
    x = "Total Plastic Collected",
    y = "Region"
  ) +
  theme_light(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

**Visual 5**

```{r}
#| label: Plastic waste vs GDP per capita
#| fig-cap: "Scatterplot showing the relationship between GDP per capita and total plastic collected across countries. Points are colored by region to highlight geographic differences."

library(scales)

joined2 %>% 
  filter(!is.na(region), !is.na(gdppercapita)) %>% 
  group_by(country, gdppercapita, region) %>% 
  summarise(
    total = sum(grand_total, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = gdppercapita, y = total, color = region)) +
  geom_point(alpha = 0.9, size = 2) +
  scale_x_log10(labels = label_dollar(scale = 1e-3, suffix = "K")) +
  scale_color_brewer(palette = "Set2", name = "Region") +
  labs(
    title = "Total Plastic Waste vs GDP per Capita",
    subtitle = "Country-level totals with regional grouping",
    x = "GDP per Capita (log scale)",
    y = "Total Plastic Collected"
  ) +
  theme_light(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
  )
```

**Visual 6**

```{r}
#| label: Top companies by region
#| fig-cap: "Bar chart showing the parent companies with the highest total recorded plastic waste, grouped by region."

top_companies_region <- joined_all %>%
  filter(
    !is.na(parent_company),
    !is.na(region),
    region != "Oceania",
    parent_company != "Grand Total",
    parent_company != "null",
    parent_company != "NULL"
  ) %>%
  group_by(region, parent_company) %>%
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(region) %>%
  slice_max(order_by = total_plastic, n = 5) %>%
  ungroup()

ggplot(top_companies_region,
       aes(x = total_plastic, y = fct_reorder(parent_company, total_plastic), fill = region)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ region, scales = "free_y") +
  labs(
    title = "Top Parent Companies by Region",
    subtitle = "Top 5 companies with the highest recorded plastic waste in each region",
    x = "Total Plastic Collected",
    y = "Parent Company"
  ) +
  theme_light(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

**Visual 7**

```{r}

```

**Statistical Model: Does GDP, Population density, and tempuertae contribute to Total Plastic Waste**

```{r}
model_data <- country_year %>%
  filter(
    total_plastic > 0,
    !is.na(gdp),
    !is.na(Temperature),
    !is.na(pop_density),
    !is.na(region)
  ) %>%
  mutate(
    log_total_plastic = log(total_plastic),
    log_gdp = log(gdp),
    log_pop_density = log(pop_density + 1)
  )

plastic_model <- lm(
  log_total_plastic ~ log_gdp + Temperature + log_pop_density  + region,
  data = model_data
)

summary(plastic_model)
```

```{r}
library(broom)
library(gt)

model_table <- tidy(plastic_model, conf.int = TRUE) %>%
  select(term, estimate, std.error, statistic, p.value, conf.low, conf.high)

model_gt <- model_table %>%
  gt() %>%
  cols_label(
    term = md("**Term**"),
    estimate = md("**Estimate**"),
    std.error = md("**Std. Error**"),
    statistic = md("**t Statistic**"),
    p.value = md("**p-value**"),
    conf.low = md("**CI Lower**"),
    conf.high = md("**CI Upper**")
  ) %>%
  fmt_number(
    columns = c(estimate, std.error, statistic, conf.low, conf.high),
    decimals = 3
  ) %>%
  fmt_number(
    columns = p.value,
    decimals = 4
  ) %>%
  tab_header(
    title = md("**Regression Model for Log Total Plastic Collected**")
  ) %>%
  tab_source_note(
    source_note = "Linear regression model predicting log total plastic collected from GDP, temperature, population density, and region."
  )

model_gt
```
