---
title: "Project 1 Checkpoint 4"
author: "Faran Igani and Tyler Stoen"
format:
  html:
    embed-resources: true
editor: visual
code-tools: true
execute:
  warning: false
  message: false
---

## Packages and Data

```{r}
#| label: Load packages 
library(tidyverse)
library(gt)
library(rvest)
library(ggplot2)
library(dplyr)
library(httr2)
library(jsonlite)
library(httr)
library(glue)
library(ggridges)
```

```{r}
#| label: Load the Data
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
```

## Github URL

<https://github.com/FaranIgani/STAT-541---Project-1>

## Checkpoint 4

-   **two** professionally formatted visualizations

    -   every visualization **must** have alternative (alt) text, non-default colors, a non-default theme, and clear labels

    -   **at least one** visualization must use a more complex geometry (e.g., `geom_segment()`, `geom_density_ridge()` `geom_ribbon()`, `geom_tile()`, `geom_hex()`)

-   fits **one** statistical model addressing one of the research questions posed

-   professionally formats the output of that model to communicate the results

    -   Formatted values (e.g., decimals, percentages, currencies)

    -   Boldface headers

    -   Captions

## Data from API

### API 1: restcountries.com

```{r}
#| label: data-from-api

base_url <- "https://restcountries.com/v4/"

restcountries <- GET(glue("{base_url}all?fields=name,continents,region,subregion,unMember,gini,density,geolocation"))

restcountries

restcountries_tidy <- restcountries %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>%
  as_tibble()

```

```{r}
#| label: merge-plastics

library(stringi)

normalize_name <- function(x) {
  x %>% 
    stri_trans_general("Latin-ASCII") %>% 
    str_to_lower() %>% 
    str_remove_all("[^a-z ]") %>% 
    str_trim() %>% 
    str_squish()
}

plastics_clean <- plastics %>% 
  filter(country != "EMPTY") %>% 
  mutate(
    country = case_when(
      country == "Korea" ~ "South Korea",
      country == "Taiwan_ Republic of China (ROC)" ~ "Taiwan",
      country == "United Kingdom of Great Britain & Northern Ireland" ~ 
        "United Kingdom",
      TRUE ~ country
    )
  ) %>%
  mutate(country_key = normalize_name(country))

restcountries_clean <- restcountries_tidy %>% 
  select(name.common, name.official, region, subregion,
         density, unMember, geolocation.latitude, 
         geolocation.longitude) %>% 
  mutate(common_key = recode(
    normalize_name(name.common),
    "ivory coast" = "cote divoire"
    ),
         official_key = normalize_name(name.official))
  

# Different name types, try both

joined <- plastics_clean %>%
  left_join(
    restcountries_clean %>% select(-official_key),
    by = c("country_key" = "common_key")
  )

matched <- joined %>% filter(!is.na(region))
unmatched <- joined %>% filter(is.na(region)) %>%
  select(names(plastics_clean))

refilled <- unmatched %>%
  left_join(
    restcountries_clean %>% select(-common_key),
    by = c("country_key" = "official_key")
  )

joined <- bind_rows(matched, refilled)
```

### API 2: worldbank

```{r}
#| label: API, World Bank  API

base_url <- "https://api.worldbank.org/v2/country/all/indicator/"

pop_code <- "SP.POP.TOTL"
gdp_code <- "NY.GDP.MKTP.CD"

pop_url <- str_c(base_url, pop_code, "?format=json&per_page=20000")
gdp_url <- str_c(base_url, gdp_code, "?format=json&per_page=20000")

pop_raw <- GET(pop_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>%
  pluck(2) %>% 
  as_tibble() %>% 
  select(
    country.value,
    year = date,
    population = value
  ) %>% 
  mutate(year = as.numeric(year)) %>% 
  filter(year %in% c(2019, 2020))

gdp_raw <- GET(gdp_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>% 
  pluck(2) %>% 
  as_tibble() %>% 
  select(
    country.value,
    year = date,
    gdp = value
  ) %>% 
  mutate(year = as.numeric(year)) %>% 
  filter(year %in% c(2019, 2020))

pop_and_gdp_raw <- inner_join(pop_raw, gdp_raw, by = c("country.value", "year")) %>% 
  mutate(country.value = recode(
           country.value,
          "Turkiye" = "Turkey",
          "Hong Kong SAR, China" = "Hong Kong",
          "Viet Nam" = "Vietnam",
          "Korea, Rep." = "South Korea",
          "United States" = "United States of America"
          ),
         country_code = normalize_name(country.value))

joined2 <- joined %>% 
  left_join(pop_and_gdp_raw, by = c("country_key" = "country_code", "year")) %>% 
  mutate(
    gdppercapita = gdp / population
  )
                              
```

## Web-Scraped Data

```{r}
#| label: Webscraping, Wiki population
#not multi page
url <- "https://en.wikipedia.org/wiki/List_of_countries_by_average_yearly_temperature"

temp_data <- read_html(url) %>%
  html_table(fill = TRUE)

temp_data <- temp_data[[1]]

names(temp_data) <- make.names(names(temp_data), unique = TRUE)

temp_data <- temp_data %>% 
  rename(country = `Countryor.region`) %>%
  mutate(
    Temperature = readr::parse_number(Temperature),
    country_key = normalize_name(country)
  ) %>% 
  select(country_key, Temperature)
```

```{r}
#| label: Webscraping, Wiki population density multi page

scrape_density <- function(country_name, wiki_name) {
  
  url <- str_c("https://en.wikipedia.org/wiki/", wiki_name)
  
  out <- tryCatch({
    page <- read_html(url)
    
    rows <- page %>%
      html_elements("table.infobox tr")
    
    row_text <- rows %>% html_text2()
    
    density_idx <- str_which(row_text, regex("density", ignore_case = TRUE))
    
    if (length(density_idx) == 0) {
      return(tibble(country_key = country_name, pop_density = NA_real_))
    }
    
    density_row <- rows[density_idx[1]]
    
    cells <- density_row %>% html_elements("th, td") %>% html_text2()
    
    if (length(cells) < 2) {
      return(tibble(country_key = country_name, pop_density = NA_real_))
    }
    
    value <- parse_number(cells[length(cells)])
    
    tibble(country_key = country_name, pop_density = value)
    
  }, error = function(e) {
    tibble(country_key = country_name, pop_density = NA_real_)
  })
  
  out
}
```

```{r}
#| label: mapping to countries in plastic data

wiki_countries <- joined2 %>%
  distinct(country_key) %>%
  mutate(
    wiki_country = recode(country_key,
      "united states of america" = "United_States",
      "hong kong" = "Hong_Kong",
      "south africa" = "South_Africa",
      "sri lanka" = "Sri_Lanka",
      "cote divoire" = "Ivory_Coast",
      "taiwan republic of china roc" = "Taiwan",
      "united kingdom of great britain northern ireland" = "United_Kingdom",
      "korea" = "South_Korea",
      .default = str_replace_all(str_to_title(country_key), " ", "_")
    )
  )

density_data <- map2_dfr(
  wiki_countries$country_key,
  wiki_countries$wiki_country,
  scrape_density
)
```

```{r}
joined_all <- joined2 %>% 
  left_join(temp_data, by = "country_key") %>% 
  left_join(density_data, by = "country_key")
```

## Visuals

```{r}
library(tidyverse)

country_year <- joined_all %>%
  group_by(country, year, region) %>%
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    gdp = mean(gdp, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    Temperature = mean(Temperature, na.rm = TRUE),
    pop_density = mean(pop_density, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(
    !is.na(total_plastic),
    !is.na(gdp),
    !is.na(Temperature),
    !is.na(region)
  )
```

**Visual 1**

```{r}
#| fig-alt: "A ridgeline density plot showing the distribution of total plastic waste per country, separated by continent. Americas has the highest peak near zero with a moderate right tail. Asia has a flatter distribution with several secondary peaks at high waste values representing individual countries producing high waste. Africa is multimodal. Europe is concentrated near zero with one outlier peak. All distributions are right-skewed."
#| fig-cap: "Density ridge plot showing the distribution of total plastic collected across regions. Each ridge represents the distribution of country-level plastic totals within a region."

vis_1 <- joined_all %>%
  filter(!is.na(region), region != "Oceania") %>% 
  group_by(country, region) %>%
  summarize(
    total = sum(grand_total, na.rm = TRUE), .groups = "drop"
  ) %>% 
  mutate(
    region = fct_reorder(region, total, .desc = TRUE)
  ) %>% 
  ggplot(aes(x=total, y=region, fill=region)) +
  geom_density_ridges(alpha = 0.5, height = 1, position = "identity") +
  labs(
    title = "Total Plastic Waste by Continent",
    subtitle = "Distribution of Total Plastic Waste by Continent (in Pieces)",
    x= "Total Plastic Waste",
    y= ""
  ) +
  scale_fill_brewer(palette = "Pastel1") +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme_minimal() +
  theme(legend.position = "none")

vis_1
```

**Visual 2**

```{r}
#| fig-alt: "A ridgeline density plot showing the distribution of plastic waste per capita in pieces per person by continent, with four ridges for Europe, Americas, Asia, and Africa. All distributions are right-skewed and peak near zero plastic items recorded per person. Europe has the widest spread with several secondary peaks and one extreme outlier. Asia also has an extreme outlier at 0.010 pieces of plastic per capita. Africa is the most concentrated near zero with a slightly wider base than the others."
#| fig-cap: "Density ridge plot showing the distribution of plastic waste per capria across regions. Each ridge represents the distribution of country-level plastic totals within a region."

vis_1.5 <- joined_all %>%
  filter(!is.na(region), region != "Oceania") %>% 
  group_by(country, region, population) %>%
  summarize(
    total = sum(grand_total, na.rm = TRUE), .groups = "drop"
  ) %>% 
  mutate(
    region = fct_reorder(region, total, .desc = TRUE),
    waste_per_capita = total/population
  ) %>% 
  ggplot(aes(x=waste_per_capita, y=region, fill=region)) +
  geom_density_ridges(alpha = 0.5, height = 1, position = "identity") +
  labs(
    title = "Plastic Waste per Capita by Continent",
    subtitle = "Distribution of Plastic Waste per Capita by Continent (in Pieces/Person)",
    x= "Plastic Waste Per Capita",
    y= ""
  ) +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal() +
  theme(legend.position = "none")

vis_1.5
```

**Visual 3**

```{r}
#| fig-alt: "A scatterplot comparing recorded plastic waste on the y-axis against GDP per capita on the x-axis, with points colored by continent. The majority of points are clustered near zero on both axes. One African country is a strong outlier with approximately 165,000 units of plastic waste recorded despite a very low GDP per capita below 5,000. Several Asian and Americas countries also show elevated waste counts at low GDP levels. European countries are spread across the full GDP per capita range but generally show low waste counts. No clear positive or negative relationship between GDP per capita and plastic waste is visible across the full dataset."
#| fig-cap: "Scatterplot showing the relationship between GDP per capita and total plastic collected across countries. Points are colored by region to highlight geographic differences."

vis_2 <- joined_all %>% 
  group_by(country, gdppercapita, region) %>% 
  summarize(
    total = sum(grand_total, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = gdppercapita, y = total, colour = region)) +
  geom_point() +
   scale_color_brewer(palette = "Set2", name = "Region") +
  labs(
    title = "Plastic Waste Per Capita",
    x= "GDP per Capita",
    y= "Total Recorded Plastic Waste"
  ) +
  ylim(c(0,180000)) +
  theme_light(base_size = 15) 

vis_2

```

**Visual 4**

```{r}
#| label: Top companies by region
#| fig-cap: "Bar chart showing the parent companies with the highest total recorded plastic waste, grouped by region."
#| fig-alt: "Faceted bar chart showing the top five parent companies contributing to plastic waste within each region (Africa, Americas, Asia, and Europe). In all regions, 'Unbranded' plastic dominates by a large margin, especially in Asia where it far exceeds other companies. Named companies such as The Coca-Cola Company, Nestle, and Unilever appear across multiple regions but contribute substantially less than unbranded sources. Overall, the distribution is highly skewed, with a single dominant category in each region and relatively small contributions from the remaining companies."

top_companies_region <- joined_all %>%
  filter(
    !is.na(parent_company),
    !is.na(region),
    region != "Oceania",
    parent_company != "Grand Total",
    parent_company != "null",
    parent_company != "NULL"
  ) %>%
  group_by(region, parent_company) %>%
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(region) %>%
  slice_max(order_by = total_plastic, n = 5) %>%
  ungroup()

ggplot(top_companies_region,
       aes(x = total_plastic, y = fct_reorder(parent_company, total_plastic), fill = region)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ region, scales = "free_y") +
  scale_x_continuous(
    breaks = seq(0, 150000, by = 50000),
    labels = scales::label_number(scale = 1e-3, suffix = "K")
  ) +
  labs(
    title = "Top Parent Companies in Plastic Waste by Region",
    subtitle = "Top 5 companies with the highest recorded plastic waste in each region",
    x = "Total Plastic Collected",
    y = ""
  ) +
  scale_fill_brewer(palette = "Dark2") +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11),
    panel.grid.minor = element_blank()
  ) +
  theme_light()
```

**Statistical Model**

```{r}
model_data <- country_year %>%
  filter(
    total_plastic > 0,
    !is.na(gdp),
    !is.na(Temperature),
    !is.na(pop_density),
    !is.na(region)
  ) %>%
  mutate(
    log_total_plastic = log(total_plastic),
    log_gdp = log(gdp),
    log_pop_density = log(pop_density + 1)
  )

plastic_model <- lm(
  log_total_plastic ~ log_gdp + Temperature + log_pop_density  + region,
  data = model_data
)

summary(plastic_model)
```

```{r}
library(broom)
library(gt)

model_table <- tidy(plastic_model, conf.int = TRUE) %>%
  select(term, estimate, std.error, statistic, p.value, conf.low, conf.high)

model_gt <- model_table %>%
  gt() %>%
  cols_label(
    term = md("**Term**"),
    estimate = md("**Estimate**"),
    std.error = md("**Std. Error**"),
    statistic = md("**t Statistic**"),
    p.value = md("**p-value**"),
    conf.low = md("**CI Lower**"),
    conf.high = md("**CI Upper**")
  ) %>%
  fmt_number(
    columns = c(estimate, std.error, statistic, conf.low, conf.high),
    decimals = 3
  ) %>%
  fmt_number(
    columns = p.value,
    decimals = 4
  ) %>%
  tab_header(
    title = md("**Regression Model for Log Total Plastic Collected**")
  ) %>%
  tab_source_note(
    source_note = "Linear regression model predicting log total plastic collected from GDP, temperature, population density, and region."
  )

model_gt
```
